%!TEX root = main.tex

We have presented how Maude has been integrated into
\Hets, a parsing, static analysis, and proof management tool that
combines various tools for different specification languages. To
achieve this integration, we consider preordered algebra semantics for
Maude and define an institution comorphism from Maude to \CASL.  This
integration allows to prove properties of Maude specifications like
those expressed in Maude views. We have also implemented a
normalization of the development graphs that allows us to prove
freeness constraints. We have used this transformation to connect
Maude to Isabelle \cite{Isabelle02}, a Higher Order Logic prover, and
have demonstrated a small example proof about reversal of lists.
Moreover, this encoding is suited for proofs of e.g.\ extensionality
of sets, which require first-order logic, going beyond the abilities
of existing Maude provers like ITP.

Since interactive proofs are often not easy to conduct, future work
will make proving more efficient by adopting automated induction
strategies like rippling~\cite{DBLP:conf/tphol/DixonF04}.  We also
have the idea to use the automatic first-order prover SPASS for
induction proofs by integrating special induction strategies directly
into \Hets.

We have also studied the possible comorphisms from \CASL to Maude. We
distinguish whether the formulas in the source theory are confluent and
terminating or not. In the first case, that we plan to check with the
Maude termination~\cite{MTT08} and confluence checker \cite{ChurchRoss10},
we map formulas to equations,
whose execution in Maude is more efficient, while in the second case
we map formulas to rules.

Finally, we also plan to relate \Hets' Modal Logic and Maude models in order to use
the Maude model checker \cite[Chapter 13]{maude-book} for linear temporal
logic.


\textbf{Acknowledgments} We wish to thank Francisco Dur\'an for discussions 
regarding freeness in Maude, Martin K\"uhl for cooperation
on the implementation of the system presented here,
and Maksym Bortin for help with the Isabelle proofs.
This work has been
supported by the German Federal Ministry of Education and Research
(Project 01 IW 07002 FormalSafe), the German Research Council (DFG)
under grant KO-2428/9-1, the Comunidad de Madrid project \emph{PROMETIDOS}
(S2009/TIC--1465), and the MICINN Spanish project
\emph{DESAFIOS10} (TIN2009-14599-C03-01).


%strat and frozen attributes. Co-algebraic constructions.