%!TEX root = main.tex

As mentioned in the introduction, Maude modules are executable rewriting logic specifications.
Rewriting logic \cite{Meseguer92-tcs} is a logic of change very suitable
for the specification of concurrent systems that is parameterized
by an underlying equational logic, for which Maude uses membership
equational logic \cite{BouhoulaJouannaudMeseguer00,Meseguer97},
which, in addition to equations, allows the statement of membership
axioms characterizing the elements of a sort. In the following sections we
present both logics and how their specifications are represented as Maude modules.

\subsection{Membership equational logic} \label{mel-section}

A \emph{signature} in membership equational logic is a triple $(K,\Sigma, S)$
(just $\Sigma$ in the following),\
with $K$ a set of {\em kinds},
$\Sigma = \{\Sigma_{k_1\ldots k_n,k}\}_{(k_1\ldots k_n,k)\in K^{*}\times K}$ a
many-kinded signature, and $S =
\{S_{k}\}_{k\in K}$ a pairwise disjoint $K$-kinded family of sets of
\emph{sorts}.
The kind of a sort $s$ is denoted by $[s]$.
We write $T_{\Sigma,k}$ and $T_{\Sigma,k}(X)$ to denote respectively the set
of ground
$\Sigma$-terms with kind $k$ and of $\Sigma$-terms with kind $k$ over variables
in $X$, where $X = \{ x_1:k_1, \dots, x_n:k_n\}$ is a set of $K$-kinded
variables.
Intuitively, terms with a kind but without a sort represent undefined or error
elements.

The atomic formulas of membership equational logic are either \emph{equations}
$t = t'$, where $t$ and $t'$ are $\Sigma$-terms of the same kind, or
\emph{membership axioms} of the form $t : s$, where the term $t$
has kind $k$ and $s \in S_k$.
\emph{Sentences} are universally-quantified Horn clauses of the
form $(\forall X)\, A_0 \Leftarrow A_1 \wedge \ldots \wedge A_n$,
where each $A_i$ is  either an equation or a membership axiom, and $X$ is a
set of $K$-kinded variables containing all the variables in the $A_i$.
A \emph{specification} is a pair $(\Sigma,E)$, where $E$ is a set
of sentences in membership equational logic over the signature $\Sigma$.

Models of membership equational logic specifications are
\emph{$\Sigma$-algebras} $\model$ consisting of a set $A_k$ for each kind $k \in K$,
a function $A_f : A_{k_1}\times \dots \times A_{k_n} \longrightarrow A_k$ for each 
operator $f \in \Sigma_{k_1 \dots k_n, k}$, and
a subset $A_s \subseteq A_k$ for each sort $s\in S_k$. The meaning
$\den{t}$ of a term $t$ in an algebra $\model$ is inductively defined as usual.
Then, an algebra $\model$ satisfies an equation $t = t'$  (or the equation holds
in the algebra), denoted $\model \models t = t'$, when both terms have the same meaning:
$\den{t} = \den{t'}$. In the same way, satisfaction of a membership is defined as:
$\model \models t : s$ when $\den{t} \in A_s$.

A membership equational logic specification $(\Sigma,E)$ has an initial model 
$\mathcal{T}_{\Sigma/E}$ whose elements
are $E$-equivalence classes of terms $[t]$.
We refer to \cite{BouhoulaJouannaudMeseguer00,Meseguer97} for a detailed presentation of
$(\Sigma,E)$-algebras,
sound and complete deduction rules, as well as the construction of
initial and free algebras.

Since the membership equational logic specifications that we consider are assumed
to satisfy the executability requirements of confluence, termination, and 
sort-decreasingness, their equations $t=t'$ can be oriented from left to right,
$t \rightarrow t'$. Such a statement holds in an algebra, denoted 
$\model \models t \rightarrow t'$, exactly when $\model \models t = t'$, i.e., when
$\den{t} = \den{t'}$. Moreover, under those assumptions an equational condition $u = v$
in a conditional equation can be checked by finding a common term $t$ such
that $u \rightarrow t$ and $v \rightarrow t$.

\subsection{Maude functional modules} \label{maudefmod}

Maude functional modules \cite[Chapter 4]{maude-book}, introduced 
with syntax \texttt{fmod ...\ endfm}, are executable membership
equational specifications and their semantics is given by the corresponding
initial membership algebra in the class of algebras satisfying the specification.

In a functional module we can declare sorts (by means of keyword
\texttt{sort}(\texttt{s})); subsort relations between sorts
(\texttt{subsort}); operators (\texttt{op}) for building values of these
sorts, giving the sorts of their arguments and result, and which may have
attributes such as being associative (\texttt{assoc}) or commutative
(\texttt{comm}), for example; memberships (\texttt{mb}) asserting that a term
has a sort; and equations (\texttt{eq}) identifying terms.  
Both memberships and equations can be conditional (\texttt{cmb} and \texttt{ceq}).

Maude does automatic kind inference from the sorts declared by the user and
their subsort relations.  Kinds are \emph{not} declared explicitly, and
correspond to the connected components of the subsort relation.
The kind corresponding to a sort \texttt{s} is denoted \texttt{[s]}.
For example, if we have sorts \texttt{Nat} for natural numbers and \texttt{NzNat} 
for nonzero natural numbers with a subsort \texttt{NzNat < Nat}, then 
\texttt{[NzNat]} = \texttt{[Nat]}.

An operator declaration like

{\codesize
\begin{verbatim}
 op _div_ : Nat NzNat -> Nat .
\end{verbatim}
}

\noindent
is logically understood as a declaration at the kind level

{\codesize
\begin{verbatim}
 op _div_ : [Nat] [Nat] -> [Nat] .
\end{verbatim}
}

\noindent
together with the conditional membership axiom

{\codesize
\begin{verbatim}
 cmb N div M : Nat if N : Nat and M : NzNat .
\end{verbatim}
}

A subsort declaration \texttt{NzNat < Nat} is logically understood as
the conditional membership axiom

{\codesize
\begin{verbatim}
 cmb N : Nat if N : NzNat .
\end{verbatim}
}

\subsection{Rewriting logic}

Rewriting logic extends equational logic by introducing the notion of \emph{rewrites} 
corresponding to
transitions between states; that is, while equations are interpreted as equalities and therefore
they are symmetric, rewrites denote changes which can be irreversible. 

A rewriting logic specification, or \emph{rewrite theory}, has the form
$\mathcal{R} = (\Sigma,E,R)$, where $(\Sigma,E)$ is an equational specification
and $R$ is a set of \emph{rules} as described below. From this definition,
one can see that rewriting logic is built on top of equational logic, so
that rewriting logic is parameterized
with respect to the version of the underlying equational logic; in our
case, Maude uses membership equational logic, as described in the
previous sections. A rule in $R$ has the general conditional
form\footnote{There is no need for the condition listing first equations,
then  memberships, and then rewrites: this is just a notational
abbreviation, they can be listed in any order.} 
\[
(\forall X) \; e \Rightarrow e' \Leftarrow \bigwedge_{i=1}^{n} u_i = u'_i \wedge
                      \bigwedge_{j=1}^{m}  v_j : s_j \wedge
                      \bigwedge_{k=1}^{l} w_k \Rightarrow w'_k
\]
where the head is a rewrite and the conditions can be equations,
memberships, and rewrites; both sides of a rewrite must have the same kind. 
From these rewrite rules, one can deduce rewrites of the form
$t \Rightarrow t'$ by means of general deduction rules introduced
in \cite{Meseguer92-tcs} (see also \cite{BruniMeseguer06}).

Models of rewrite theories are called \emph{$\mathcal{R}$-systems}
in \cite{Meseguer92-tcs}.
Such systems are defined as categories that possess a
$(\Sigma,E)$-algebra structure, together with a natural transformation
for each rule in the set $R$. More intuitively, the idea is that we have a
$(\Sigma,E)$-algebra, as described in Section~\ref{mel-section}, with
transitions between the elements in each set $A_k$; moreover, these
transitions must satisfy several additional requirements, including that
there are identity transitions for each element, that transitions can
be sequentially composed, that the operations in the signature $\Sigma$
are also appropriately defined for the transitions, and that we have
enough transitions corresponding to the rules in $R$. Then, if we keep in
this context the notation $\model$ to denote an $\mathcal{R}$-system, a
rewrite $t \Rightarrow t'$ is satisfied by $\model$, 
denoted $\model \models t \Rightarrow t'$, when there is a transition
$\den{t} \rightarrow_\model \den{t'}$ in the system between the
corresponding meanings of both sides of the rewrite, where $\rightarrow_\model$
will be our notation for such transitions. 

The rewriting logic deduction rules introduced in \cite{Meseguer92-tcs}
are sound and complete with respect to this notion of model. Moreover,
they can be used to build initial and free models; see \cite{Meseguer92-tcs}
for details.

\subsection{Maude system modules}

Maude system modules \cite[Chapter 6]{maude-book}, introduced with
syntax \texttt{mod ...\ endm}, are executable rewrite 
theories and their semantics is given by the initial system in the class of 
systems corresponding to the rewrite theory.  A system module can contain all the
declarations of a functional module and, in addition, declarations for
rules (\texttt{rl}) and conditional rules (\texttt{crl}).

The executability requirements for equations and memberships in a system
module are the same as those of functional modules, namely, confluence,
termination, and sort-decreasingness. With respect to rules, the satisfaction
of all the conditions in a conditional rewrite rule is attempted sequentially 
from left to right, solving rewrite conditions by means of search; 
for this reason, we can have new variables in such conditions but they
must become instantiated along this process of solving from left to right
(see \cite{maude-book} for details). Furthermore, the strategy followed
by Maude in rewriting with rules is to compute the normal form of a term
with respect to the equations before applying a rule. This strategy is
guaranteed not to miss any rewrites when the rules are \emph{coherent}
with respect to the equations \cite{eq-rl-rwl,maude-book}. In a way
quite analogous to confluence, this coherence requirement means that, given
a term $t$, for each rewrite of it using a rule in $R$ to some term $t'$,
if $u$ is the normal form of $t$ with respect to the equations and
memberships in $E$, then there is a rewrite of $u$ with some rule in
$R$ to a term $u'$ such that $u' =_E t'$ (that is, the equation $t' = u'$
can be deduced from $E$).

\subsection{Advanced features}\label{subsec:adv_feat}

In addition to the modules presented thus far, we present in this section
some other Maude features that will be used throughout this paper.

\subsubsection{Module operations}

To ease the specification of large systems, Maude provides several
mechanisms to structure them. We describe in this section these
procedures, that will be used later to build the development graphs.

Maude modules can import other modules in three different modes:
\begin{itemize}
\item
The \verb"protecting" mode (abbreviated as \verb"pr") indicates that \emph{no
junk and no confusion} can be added to the module, where junk refers to
new terms in canonical form while confusion implies that different canonical
terms in the initial module are made equal by equations in the imported module.

\item
The \verb"extending" mode (abbreviated as \verb"ex") indicates that junk is
allowed but confusion is forbidden.

\item
The \verb"including" mode (abbreviated as \verb"inc") allows both
junk and confusion.

\end{itemize}

More specifically, these importation modes do not import modules but
\emph{module expressions} that, in addition to a single module identifier,
can be:

\begin{itemize}
\item
A summation of two module expressions $\mathit{ME}_1 \,\verb"+"\,\mathit{ME}_2$,
which creates a new module that includes all the information in its summands.

\item
A renaming $\mathit{ME} \,\verb"*"\, \verb"("\mathit{Renaming}\verb")"$, where 
$\mathit{Renaming}$ is a list of renamings. They can be renaming of sorts;

$$
\verb"sort"\; \mathit{sort}_1\; \verb"to"\; \mathit{sort}_2\; \verb"."
$$

of operators, distinguishing whether it renames all the operators
with the given identifier (when the attributes are modified, only
\verb"prec", \verb"gather", and \verb"format" are allowed)

$$
\verb"op"\; \mathit{id}_1\; \verb"to"\; \mathit{id}_2\; \verb"."
$$
\vspace{-3ex}
$$
\verb"op"\; \mathit{id}_1\; \verb"to"\; \mathit{id}_2\; \verb"[" \mathit{atts} \verb"]"
\;  \verb"."
$$

\noindent or it renames the operators of the given arity;

$$
\verb"op"\; \mathit{id}_1 : \mathit{arity}\; \verb"->" \; \mathit{coarity}\;
\verb"to"\; \mathit{id}_2\; \verb"."
$$
\vspace{-3ex}
$$
\verb"op"\; \mathit{id}_1 : \mathit{arity}\; \verb"->" \; \mathit{coarity}\;
\verb"to"\; \mathit{id}_2\; \verb"[" \mathit{atts} \verb"]" \; \verb"."
$$

or of labels:

$$
\verb"label"\; \mathit{label}_1\; \verb"to"\; \mathit{label}_2\; \verb"."
$$

\end{itemize}

\subsubsection{Theories}\label{subsec:theories}

Theories are used to declare module interfaces, namely the syntactic
and semantic properties to be satisfied by the actual parameter modules
used in an instantiation. As for modules, Maude supports two different types
of theories: functional theories and system theories, with the same structure
of their module counterparts, but with a different semantics. Functional
theories  are declared with the keywords \verb"fth ... endfth", and
system theories with the keywords \verb"th ... endth". Both of them can
have sorts, subsort relationships, operators, variables, membership axioms,
and equations, and can import other theories or modules. System theories can
also have rules. Although there is no restriction on the operator attributes
that can be used in a theory, there are some subtle restrictions and
issues regarding the mapping of such operators (see Section
\ref{subsec:views}).

Like functional modules, functional theories are membership equational
logic theories, but they do not need to be Church-Rosser and terminating,
and  therefore some or all of their statements may be declared with the
\verb"nonexec" attribute and can only be executed in a controlled way.

\subsubsection{Views}\label{subsec:views}

We use views to specify how a particular target module or theory
satisfies a source theory. In general, there may be several ways in which 
such requirements might be satisfied by the target module or
theory; that is, there can be many different views, each specifying a
particular interpretation of the source theory in the target. Each
view declaration has an associated set of proof obligations, namely, for
each axiom in the source theory it should be the case that the axiom's
translation by the view holds true in the target. Since the target can
be a module interpreted initially, verifying such proof obligations may
in general require inductive proof techniques. Such proof obligations
are not discharged or checked by the system. 
In the definition of a view we have to indicate its name, the source
theory, the target module or theory, and the mapping of each sort and
operator in the source theory. The source and target of a
view can be any module expression, with the source module expression
evaluating to a theory and the target module expression evaluating to a
module or a theory.

The mappings allowed in views are:

\begin{itemize}

\item
Mappings between sorts:

$$
\verb"sort"\; \mathit{sort}_1\; \verb"to"\; \mathit{sort}_2\; \verb"."
$$

\item
Mappings between operators, where the user can specify the arity and coarity
of the operator to disambiguate them:

$$
\verb"op"\; \mathit{id}_1\; \verb"to"\; \mathit{id}_2\; \verb"."
$$
\vspace{-3ex}
$$
\verb"op"\; \mathit{id}_1 : \mathit{arity}\; \verb"->" \; \mathit{coarity}\;
\verb"to"\; \mathit{id}_2\; \verb"."
$$

\item
In addition to these mappings, the user can map a term $\mathit{term}_1$, that
can only be a single operator applied to variables, to any term $\mathit{term}_2$
in the target module, where the sort of the variables in the first term has been
translated by using the sort mappings. Note that in that case the arity of the
operator in the source theory and the one in the target module can be different:

$$
\verb"op"\; \mathit{term}_1\; \verb"to term"\; \mathit{term}_2\; .
$$

\end{itemize}

Notice that we cannot map labels, and thus we cannot identify the statements in
the theory with those in the target module.

\subsubsection{Parameterized modules}\label{subsec:pmod}

Maude modules can be parameterized. A parameterized
system module has syntax

$$
\verb"mod M{" X_1 :: T_1 , \ldots , X_n :: T_n \verb"} is ... endm"
$$

\noindent with $n \geq 1$. Parameterized functional modules have completely
analogous syntax.

The \verb"{"$X_1 :: T_1 , \ldots , X_n :: T_n$\verb"}" part is called the
interface, where each pair $X_i :: T_i$ is a parameter, and each $X_i$ is an
identifier---the parameter name or parameter label---and each $T_i$ is
an expression that yields a theory---the parameter theory. Each parameter
name in an interface must be unique, although there is no uniqueness
restriction on the parameter theories of a module. The parameter theories
of a functional module must be functional theories.

In a parameterized module $M$, all the sorts and statement labels
coming from theories in its interface must be qualified by their names. Thus,
given a parameter $X_i :: T_i$, each sort $S$ in $T_i$ must be
qualified as $X_i\texttt{\$}S$, and each label $l$ of a statement occurring in
$T_i$ must be qualified as $X_i\texttt{\$}l$. In fact, the parameterized module
$M$ is flattened as follows. For each parameter $X_i :: T_i$, 
a renamed copy of the theory $T_i$, called $X_i :: T_i$ is included.
The renaming  maps each sort $S$ to $X_i\texttt{\$}S$, and each label $l$
of a statement occurring in $T_i$ to $X_i\texttt{\$}l$. The renaming has
no effect on importations of modules. Thus, if $T_i$ includes a theory $T'$,
when the renamed
theory $X_i :: T_i$ is created and included into $M$, and the renamed
theory $X_i :: T'$ will also be created and included into $X_i :: T_i$. 
However, the renaming will have no effect on modules imported by either the
$T_i$ or $T'$; for example, if \verb"BOOL" is imported by one of these
theories, it is not renamed, but imported in the same way into $M$.
%
Moreover, sorts declared in parameterized modules can also be parameterized,
and these may duplicate, omit, or reorder parameters.

The parameters in parameterized modules are bound to the formal parameters
by \emph{instantiation}. The 
instantiation requires a view from each formal parameter to its corresponding 
actual parameter. Each such view is then used to bind the names of sorts, 
operators, etc. in the formal parameters to the corresponding sorts, operators 
(or expressions), etc. in the actual target.
The instantiation of a parameterized module must be made with views 
explicitly defined previously.

















